---
title: "Getting Started with ssfaitk"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ssfaitk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to TRUE when you have the package installed
)
```

## Introduction

The SSF AI Toolkit (`ssfaitk`) provides machine learning models for analyzing small-scale fisheries data. This vignette demonstrates the basic workflow for:

1. Setting up the package
2. Predicting fishing effort from GPS tracks
3. Predicting gear types from trip summaries
4. Predicting vessel types from trip summaries

## Installation and Setup

### Install the R Package

```{r install}
# Install from GitHub
devtools::install_github("WorldFishCenter/ssf-ai-toolkit", subdir = "Rplug")
```

### Load the Package

```{r load}
library(ssfaitk)
```

### Install Python Dependencies

The R package requires the Python `ssfaitk` package. Install it using:

```{r install_python}
# For local development (if you have the source code)
install_ssfaitk()

# For remote installation from GitHub
install_ssfaitk(
  package_url = "git+https://github.com/WorldFishCenter/ssf-ai-toolkit.git"
)
```

### Verify Installation

```{r verify}
# Check if Python package is available
ssfaitk_available()  # Should return TRUE

# Display Python environment information
check_python_env()
```

## Fishing Effort Classification

Fishing effort classification identifies fishing vs. non-fishing activity from GPS tracking data.

### Sample Data

Let's create some sample GPS tracking data:

```{r sample_tracks}
# Create sample GPS tracks
set.seed(123)
n_points <- 100

sample_tracks <- data.frame(
  trip_id = rep(1:5, each = 20),
  timestamp = seq(as.POSIXct("2024-01-01 06:00:00"),
                  by = "5 min",
                  length.out = n_points),
  latitude = -6.0 + cumsum(rnorm(n_points, 0, 0.001)),
  longitude = 39.0 + cumsum(rnorm(n_points, 0, 0.001))
)

head(sample_tracks)
```

### Option 1: Statistical Prediction (No Training Required)

The statistical classifier uses behavioral rules and doesn't require a trained model:

```{r statistical_effort}
# Predict fishing effort using rule-based classifier
predictions <- effort_predict_statistical(sample_tracks)

# View predictions
head(predictions[, c("trip_id", "is_fishing", "fishing_score", "activity_type")])

# Summary
table(predictions$is_fishing)
table(predictions$activity_type)
```

The statistical classifier returns several useful columns:

- `is_fishing`: Binary prediction (0 = non-fishing, 1 = fishing)
- `fishing_score`: Continuous score (0-1)
- `activity_type`: Classified activity (fishing, sailing, etc.)
- `trip_phase`: Trip phase (starting, in_progress, ending)
- Additional features (speed, acceleration, turning angles, etc.)

### Option 2: Machine Learning Prediction (Requires Training)

For ML-based prediction, you need to train a model first:

```{r ml_effort}
# Create labeled training data
training_data <- sample_tracks
training_data$Activity <- sample(
  c("Fishing", "Searching", "Sailing", "Traveling"),
  nrow(training_data),
  replace = TRUE,
  prob = c(0.3, 0.2, 0.3, 0.2)
)

# Train a model
model <- effort_fit(
  training_data,
  label_col = "Activity",
  save_path = "effort_model.joblib"
)

# Predict on new data
new_tracks <- sample_tracks  # In practice, this would be unlabeled data
predictions <- effort_predict(
  new_tracks,
  model_path = "effort_model.joblib"
)

# View predictions
head(predictions[, c("trip_id", "effort_pred", "effort_prob")])
```

### Custom Column Names

If your data uses different column names:

```{r custom_columns}
# Data with custom column names
custom_tracks <- sample_tracks
colnames(custom_tracks) <- c("VesselID", "GPS_Time", "Lat", "Lon")

# Use colmap to specify column mapping
predictions <- effort_predict_statistical(
  custom_tracks,
  colmap = list(
    trip_id = "VesselID",
    time = "GPS_Time",
    lat = "Lat",
    lon = "Lon"
  )
)
```

## Gear Type Prediction

Gear prediction classifies the type of fishing gear used (e.g., gillnet, handline, longline).

### Sample Trip-Level Data

```{r sample_trips}
# Create sample trip-level features
sample_trips <- data.frame(
  trip_id = 1:50,
  duration_hours = runif(50, 2, 8),
  distance_nm = runif(50, 5, 25),
  mean_sog = runif(50, 1.5, 4.5)
)

head(sample_trips)
```

### Train a Gear Model

```{r gear_training}
# Add gear labels for training
training_trips <- sample_trips
training_trips$gear_label <- sample(
  c("gillnet", "handline", "longline"),
  nrow(training_trips),
  replace = TRUE
)

# Train model
gear_model <- gear_fit(
  training_trips,
  label_col = "gear_label",
  save_path = "gear_model.joblib"
)
```

### Predict Gear Types

```{r gear_prediction}
# Predict on new trips
new_trips <- sample_trips
predictions <- gear_predict(
  new_trips,
  model_path = "gear_model.joblib"
)

# View results
table(predictions$gear_pred)
head(predictions[, c("trip_id", "duration_hours", "gear_pred")])
```

## Vessel Type Prediction

Vessel type prediction classifies vessels as motorized or non-motorized.

### Train a Vessel Model

```{r vessel_training}
# Add vessel type labels
training_trips$vessel_type_label <- sample(
  c("motorized", "non-motorized"),
  nrow(training_trips),
  replace = TRUE,
  prob = c(0.7, 0.3)
)

# Train model
vessel_model <- vessel_fit(
  training_trips,
  label_col = "vessel_type_label",
  save_path = "vessel_model.joblib"
)
```

### Predict Vessel Types

```{r vessel_prediction}
# Predict on new trips
predictions <- vessel_predict(
  new_trips,
  model_path = "vessel_model.joblib"
)

# View results
table(predictions$vessel_type_pred)
```

## Data Requirements Summary

### Effort Classification (GPS Tracks)

**Required columns** (auto-detected):

- Trip ID: `trip_id`, `Trip_ID`, `TRIP_ID`, `tripid`
- Timestamp: `timestamp`, `time`, `ltime`, `datetime`
- Latitude: `latitude`, `lat`, `LAT`, `y`
- Longitude: `longitude`, `lon`, `LON`, `x`

**Data level**: Point-by-point GPS coordinates

### Gear & Vessel Prediction (Trip Features)

**Required columns** (exact names):

- `duration_hours`: Trip duration in hours
- `distance_nm`: Total distance in nautical miles
- `mean_sog`: Mean speed over ground

**Data level**: One row per trip (aggregated from GPS points)

## Working with Real Data

### Loading Data

```{r real_data}
# Load GPS tracks from CSV
tracks <- read.csv("path/to/your/gps_data.csv")

# Load trip features
trips <- read.csv("path/to/your/trip_features.csv")
```

### Data Preparation

```{r data_prep}
# Ensure timestamp is in proper format
tracks$timestamp <- as.POSIXct(tracks$timestamp)

# Check for required columns
required_cols <- c("trip_id", "timestamp", "latitude", "longitude")
stopifnot(all(required_cols %in% colnames(tracks)))

# Check data quality
summary(tracks)
```

### Handling Large Datasets

For large datasets, consider processing in chunks:

```{r chunking}
# Process tracks in chunks
chunk_size <- 50000
n_chunks <- ceiling(nrow(large_tracks) / chunk_size)

predictions_list <- list()

for (i in 1:n_chunks) {
  start_idx <- (i - 1) * chunk_size + 1
  end_idx <- min(i * chunk_size, nrow(large_tracks))

  chunk <- large_tracks[start_idx:end_idx, ]
  predictions_list[[i]] <- effort_predict_statistical(chunk)

  message("Processed chunk ", i, " of ", n_chunks)
}

# Combine results
all_predictions <- do.call(rbind, predictions_list)
```

## Visualization

### Fishing Activity Over Time

```{r viz_time, eval=FALSE}
library(ggplot2)
library(dplyr)

# Aggregate fishing activity by hour
hourly_activity <- predictions %>%
  mutate(hour = as.POSIXct(format(timestamp, "%Y-%m-%d %H:00:00"))) %>%
  group_by(hour) %>%
  summarize(
    n_fishing = sum(is_fishing == 1),
    n_total = n(),
    pct_fishing = n_fishing / n_total * 100
  )

ggplot(hourly_activity, aes(x = hour, y = pct_fishing)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Fishing Activity Over Time",
    x = "Time",
    y = "% of Points Classified as Fishing"
  ) +
  theme_minimal()
```

### Gear Type Distribution

```{r viz_gear, eval=FALSE}
gear_summary <- predictions %>%
  count(gear_pred) %>%
  mutate(percentage = n / sum(n) * 100)

ggplot(gear_summary, aes(x = gear_pred, y = percentage, fill = gear_pred)) +
  geom_col() +
  labs(
    title = "Distribution of Gear Types",
    x = "Gear Type",
    y = "Percentage of Trips"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Troubleshooting

### Python Package Not Found

```{r troubleshoot1}
# Check if Python is configured
library(reticulate)
py_config()

# Check if ssfaitk is available
ssfaitk_available()

# If not, install it
install_ssfaitk()
```

### Column Name Issues

```{r troubleshoot2}
# Check your actual column names
colnames(your_data)

# Use colmap to map custom names
predictions <- effort_predict(
  your_data,
  model_path = "model.joblib",
  colmap = list(
    trip_id = "your_trip_column_name",
    time = "your_time_column_name",
    lat = "your_latitude_column_name",
    lon = "your_longitude_column_name"
  )
)
```

## Next Steps

- **Effort Classification**: See `vignette("effort-classification")` for advanced techniques
- **Gear Prediction**: See `vignette("gear-prediction")` for detailed workflow
- **Vessel Prediction**: See `vignette("vessel-prediction")` for vessel classification
- **Remote Deployment**: See `vignette("remote-deployment")` for using in GitHub Actions/Docker

## Getting Help

- **Function documentation**: Use `?function_name` (e.g., `?effort_predict`)
- **GitHub Issues**: https://github.com/WorldFishCenter/ssf-ai-toolkit/issues
- **Discussions**: https://github.com/WorldFishCenter/ssf-ai-toolkit/discussions
